From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-20 thesis/paper.tex
more work

\documentclass[a4paper]{article}

\usepackage[english]{babel}

\usepackage{amssymb,bbm,qtree,graphicx,natbib,aleks,linguex}

%% note that this file depends on being processed with XeTeX
\usepackage{xltxtra,xunicode,fontspec}

% font specifications (don't overuse ;-)
\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\setromanfont [Ligatures={Common,Rare}]{Gentium}

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}{\textsf} % introduce a term
\newcommand{\code}{\texttt} % layout of in-line code snippets
\newcommand{\pn}{\textsc} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear

@@

%theorem environments
%\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
%better title fixme

@@

% FIXME: don't really like this footnote
\footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
of the input and \emph{informativity} denotes \emph{validity}. Thus, if
\Disc, the discourse so far and $φ$ a new input token (fully parsed
sentence), then consistency means $\Disc → \phi$ holds and informativity
means $\Disc \nvDash$ φ.}

@@

@@

\begin{quote}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain features\footnote{To quote \cite{gs:q}: \quote{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \emph{interrogative}. Finally, a 
  \term{question} will denote the semantic entity uttered by the
  \emph{interrogative act}, whichever form it may have.
\end{quote}

@@

While there exist inferencing tools for modal logic\footnote{See %todo todo and todo 
for discussions. The interested reader is also referred to \pn{Molle}
(\url{http://molle.sf.net}), a modal theorem prover that comes with a graphical user
interface that allows interactive investigation of the constructed indices.},
a restructuring of the \pn{Curt} system to take advantage of them would be
beyond the scope of this thesis.

It was therefore decided to avoid modal logic and implement a pragmatic approach
to representing the indices of higher-order logic in \prol in section
\ref{sec:indices}.

@@

Question Semantics is a much-disputed but often overlooked field of modern
semantic research. Often question semantics has been reduced to an issue of
pragmatics, treating direct questions (interrogatives) as nothing more than a
paraphrase of indirect questions (embedded interrogatives).\cite{tichy}
even goes as far as claiming equality in semantic value between
indicatives and interrogatives. The benefit of such an approach would be the
strict reliance on traditional analyses, where interrogatives would only infer a
different `attitude' in the speaker's utterance. Putting this attitude into
context and making it yield the correct results would then be left to
pragmatics.
This approach has been challenged and proven wrong by \cite{gs:q},
who also argue that the proper treatment of questions would require extensions
to the existing logic frameworks used in natural language semantics semantics.
More recent publications, such as \cite{gal:tmbi}
address this topic and aim to introduce notions of \term{information change
potential} as opposed to the traditional focus of natural language semantics on
\term{truth values}.

Nevertheless, progress has been very lively since the first
formal approaches to the field started with \cite{hamblin:q} and other
publications such as % todo
. \cite{gs:q} provide a very descriptive overview of the field of question
semantics

@@

Our aim is to improve coverage of grammatical structures and thus when faced
with the necessity to make compromise about the complexity of our grammar, we
will choose overgeneration in favor of unmotivated complexity of the grammar
itself.

@@

\subsubsection{Enhancing Storage}

\curt parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller storage}. That name will be used from here % FIXME sucks. Please change
to refer to this particular implementation as presented in
\cite{blackburnbos:cl1}. Although \pn{Keller storage} is already quite capable
of dealing with more complex constituents such as nested noun phrases, it needs
some enhancements in order to be able to treat verbs embedding whole
sentences or \wh-phrases correctly. The aim is to enable \curt's implementation of
\pn{Keller storage} to deal with examples such as \ref{ex:kel1}\footnotemark.

\ex. \label{ex:kel1} Mia knows whether Vincent snorts.\\
\Tree [.t [.T Mia ] [.IV [.{IV/$\bar{t}$} knows ] [.{$\bar{t}$} whether [.t [.T
Vincent ] [.IV snorts ] ] ] ] ]

>>>>

\documentclass[12pt,a4paper]{article}

%\usepackage[english]{babel}

\usepackage{amssymb,qtree,natbib,aleks,linguex}

\usepackage{amsthm}

%% note that this file depends on being processed with XeTeX
\usepackage{fontspec}
\usepackage{xltxtra,xunicode}

\usepackage{mathpazo}
% font specifications (don't overuse ;-)
\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\setromanfont [Ligatures={Common,Rare}, BoldFont={Fontin Bold}, ItalicFont={Fontin Italic}]{Fontin Regular}
\setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}{\textsf} % introduce a term
\newcommand{\code}{\texttt} % layout of in-line code snippets
\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear
\newcommand{\example}{\textit} % in-text examples

@@

%theorem environments
\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
%better title fixme

@@

% FIXME: don't really like this footnote
\footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
of the input and \emph{informativity} denotes \emph{validity}. Thus, if
\Disc, the discourse so far and $\phi$ a new input token (fully parsed
sentence), then consistency means $\Disc \to \phi$ holds and informativity
means $\Disc \nvDash \phi$.}

@@

This section first gives an overview over the terminology that will be used for
the discussion of the formal treatment of questions. It then elaborates on the
differences between different logical formalisms employed in the treatment of
natural language questions. After a small overview on the types of questions
(and pointing out the subset of questions that will be treated), %fixme maybe a
% bit sloppy
the discussion
turns to how those questions are best treated in the given formalism.

The reader should note that all choices presented in this section have been made
with the implementation of section \label{sec:implementation} in mind.
% todo talk a bit more

@@

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain features\footnote{To quote \cite{gs:q}: \quote{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \emph{interrogative}. Finally, a 
  \term{question} will denote the semantic entity uttered by the
  \emph{interrogative act}, whichever form it may have.
\end{termin}

@@

While there exist inferencing tools for modal logic\footnote{See %todo todo and todo 
for discussions. The interested reader is also referred to \pn{Molle}
(\url{http://molle.sf.net}), a modal theorem prover that comes with a graphical user
interface which allows interactive investigation of the constructed indices.},
a restructuring of the \pn{Curt} system to take advantage of them would be
beyond the scope of this thesis.

It was therefore decided to avoid modal logic and implement a pragmatic approach
to representing the indices of higher-order logic in \prol in section
\ref{sec:indices}. This logic however has to be augmented with question
operators as already used by \cite{karttunen:1977}. % todo: writeme!

@@

Question semantics are an interesting subproblem of modern research on natural
language semantics. While traditional approaches based on \abbr{IL} and newer
accounts also employ plain extensional logic, % todo: which ones 
the treatment
of questions seems to demand that we introduce a new kind of logic formalism.
But as interesting topics go, the semantics of questions has attracted many
different views on this issue.

Often question semantics has been reduced to a problem of
pragmatics, treating direct interrogatives as nothing more than a
paraphrase of embedded interrogatives.\cite{tichy}
even goes as far as claiming equality in semantic (truth) value between
indicatives and interrogatives. The benefit of such an approach would be the
strict reliance on traditional analyses, where interrogatives would only infer a
different `attitude' in the speaker's utterance. Putting this attitude into
context and making it yield the correct results would then be left to
pragmatics.

This approach has been challenged and proven unfeasible by \cite{gs:q},
who have also proposed interesting extensions
to the existing logic frameworks used in natural language semantics.
More recent publications, such as \cite{gal:tmbi}
address this topic and aim to introduce notions of \term{information change
potential} as opposed to the traditional focus of natural language semantics on
\term{truth values}.

The present work will rely on \cite{gs:q} for providing the main guidelines for
the theoretic aspects and also lend from earlier works, most importantly
\cite{gs:sqpa}\footnotemark. Later on, in section \ref{sec:altq} we will also
discuss the very newest (at the time of writing) developments in question
semantics, depicted in \cite{g:is}.

\footnotetext{\cite{gs:sqpa} is actually a series of three articles
that have appeared before. Among them, we will mostly consider \cite{gs:sawhq}.
Both, the earlier paper and the later dissertation (and many other papers
originating in Amsterdam) have been kindly made
available electronically, free of charge, by the Universiteit van Amsterdam at
the following address: \url{http://dare.uva.nl}}

@@

The present work will not consider a full analysis of \wh-embedding verbs, but
confine itself on a few ones, namely \example{know}, \example{think} and
\example{believe}. The grammar implementation of \acurt allows for \wh-embedding
verbs to have a \code{type} field, determining their semantic denotation. The two
types used so far are \code{assert} for assertive verbs such as \example{know}
and \code{stipul} for stipulating verbs, such as \example{believe} and
\example{think}.

\subsubsection{Grammatic Constraints}

With this ontology in mind, the next step is to analyse and enumerate the sorts
of grammatic constructions that are to be treated. The following examples will
illustrate them.

\begin{itemize}
  \item Alternative Questions
  \ex. \a. Mia \emph{knows} whether Vincent snorts.
  \b. * Mia \emph{believes} whether Vincent snorts.
  \b. * Mia \emph{thinks} whether Vincent snorts.

  \item Enumerative Questions
  \ex. \a.  Mia \emph{knows} who snorts.
  \b. * Mia \emph{believes} who snorts.
  \b. * Mia \emph{thinks} who snorts.

  \item Embedding Propositions
  \ex. \label{ex:stipul}
  \a. Mia \emph{believes} Vincent snorts.
  \b. * Mia \emph{believes} who snorts.

\end{itemize}

An important observation to be made here is that only \emph{assertive} verbs can
embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
in \ref{ex:stipul}.
Nevertheless, both are treated in the following analysis, but with a focus on
the treatment of assertive \wh-embedding verbs.

@@

The underlying theory for

\subsubsection{Defining the Grammar Rules}

\curt parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller storage}. That name will be used from here % FIXME sucks. Please change
to refer to this particular implementation as presented in
\cite{blackburnbos:cl1}.

\pn{Keller storage} is already quite capable
of dealing with more complex constituents such as nested noun phrases, but it
has its limitations. In particular, it can only handle quantifier scope
ambiguities and does not take negation into account. Moreover, it overgenerates
with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
sentences such as \ref{ex:kellersucks}, where Keller storage produces six
instead of the desired five readings.
\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}

Clearly, \example{a hash bar} cannot take scope over \example{every owner}, but
Keller storage predicts this reading.}.

In order to cope with more complicated quantifier configurations, where
constraints need to be enforced, and with other kinds of ambiguities,
Blackburn and Bos present an approach utilizing underspecified representations and
dominance relations, \pn{Hole Semantics}\footnote{Also presented in % todo:cite
%Bos' paper
}. Since, however, none of the cases discussed so far would require the enhanced
expressive power of \pn{Hole Semantics} it was decided to go on using \pn{Keller
Storage} instead, since this is what the original implementation of the \curt
system relies on.
%todo: fix capitalization in Keller storage all over the document

Enhancing the grammatic rules of the \curt system to cope with
proposition-embedding verbs is made trivial by its elaborate grammatical back
end. \ref{ex:kel1} is an example of the logic behind the grammar that is put to
use.

\ex. \label{ex:kel1} Mia knows whether Vincent snorts.\\
\Tree
[.$t$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$ [.{$IV/\bar{t}$} { $\lambda X.know(X)$\\knows } ]
[.{$\bar{t}$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ] ]
